Efficient inference using the approximate Riemannian geometry of stochastic population models

### 2012-08-02 paper/paper.tex
Added to introduction.

\title[Fast approximate proposals for exact inference in stochastic kinetic
models]{Fast approximate proposals for exact inference in stochastic kinetic
  models}

\author{B. Calderhead$^{*}$\email{email@address.com} \\
	   Institute of Physics, CA 560034, USA
	   \and 
	   C. S. Gillespie$^{*}$\email{colin.gillespie@newcastle.ac.uk}\\
	   School of Mathematics \& Statistics, Newcastle University,
	   NE1 7RU, UK
	   }

@@

\begin{abstract}
\begin{itemize}
\item Inference for MJP is difficult: non-linear likelihood, computational
  expensive, proposals expensive, analytically intractable likelihood
\item Data: Partially observed discrete, noisy time-course.
\item Existing work for likelihood evaluations: 
\begin{itemize}
\item Likelihood calculations
\begin{itemize}
\item Exact: Boys et al, slow, not noise
\item Approx SDE - Andy and Darren (with Noise)
\item Approx MC - Andy and Colin (no noise)
\item Approx LNA - Mark (no noise), Kolmo' (no noise, but check)
\end{itemize}
\item New:
\begin{itemize}
\item Approximate proposal mechanism based on the geometry of MC
\begin{itemize}
\item Overcome strong correlation structure which is standard with partially
  observed data
\item Fast proposals based on approximate geometry
\item Accept/reject with approximate model before trying exact model
\end{itemize}
\item Inference is exact
\item Including observational error
\item Time course measurements, unlike Mark's LNA
\end{itemize}
\item Examples:
\begin{itemize}
\item Immigration-death model: exact proposal compared with approximate proposal
  (no noise)
\item Aphid model: 3 blocks * 3 water treats * 3 nitro treats with Poisson error
\item Hidden states (counts), updated using standard method
\end{itemize}
\end{itemize}
\end{itemize}
\end{abstract}

@@

@@

\subsection{Stochastic kinetic models}

\subsection{Moment closure scheme}

@@

\subsection{Delay acceptance scheme}

>>>>

\title[Efficient inference using the approximate Riemannian geometry of stochastic population models]{Efficient inference using the approximate Riemannian geometry of stochastic population models}

\author{B. Calderhead$^{*}$\email{b.calderhead@ucl.ac.uk} \\
	   Centre for Mathematics and Physics in the Life Sciences and Experimental Biology,\\University College London, UK
	   \and 
	   C. S. Gillespie$^{*}$\email{colin.gillespie@newcastle.ac.uk}\\
	   School of Mathematics \& Statistics, Newcastle University,
	   NE1 7RU, UK
	   }

@@

\begin{abstract}
\begin{itemize}
\item We present efficient and novel methodology for Bayesian inference over stochastic kinetic models with complex dynamics and large numbers of parameters.

Inference for MJP is difficult: non-linear likelihood, computational
  expensive, proposals expensive, analytically intractable likelihood
\item Data: Partially observed discrete, noisy time-course.
\item Existing work for likelihood evaluations: 
\begin{itemize}
\item Likelihood calculations
\begin{itemize}
\item Exact: Boys et al, slow, not noise
\item Approx SDE - Andy and Darren (with Noise)
\item Approx MC - Andy and Colin (no noise)
\item Approx LNA - Mark (no noise), Kolmo' (no noise, but check)
\end{itemize}
\item New:
\begin{itemize}
\item Approximate proposal mechanism based on the geometry of MC
\begin{itemize}
\item Overcome strong correlation structure which is standard with partially
  observed data
\item Fast proposals based on approximate geometry
\item Accept/reject with approximate model before trying exact model
\end{itemize}
\item Inference is exact
\item Including observational error
\item Time course measurements, unlike Mark's LNA
\end{itemize}
\item Examples:
\begin{itemize}
\item Immigration-death model: exact proposal compared with approximate proposal
  (no noise)
\item Aphid model: 3 blocks * 3 water treats * 3 nitro treats with Poisson error
\item Hidden states (counts), updated using standard method
\end{itemize}
\end{itemize}
\end{itemize}
\end{abstract}

@@

Importance of stochastic population models for further scientific understanding of natural systems vs deterministic models.  For example in ecology, where we often have to deal with discrete noisy data.  Importance of parametric sensitivity has already been shown for determining vital properties of biological systems such as robustness and also in experimental design where the sensitivity gives information about the identifiability of individual rate constants.

The sensitivity of parameters is also of central importance when inferring unknown parameter values for a mathematical model

Performing exact inference over stochastic kinetic models described by Markov Jump Processes is extremely difficult.  First and foremost, the likelihood is usually analytically intractable for models with levels of complexity suitable for describing natural phenomena.

Approximations of the likelihood surface, such as LNA and moment closure.  Parameter inference - want to explore the likelihood surface - can consider it as a Riemannian manifold.  This manifold is an approximation of the true underlying Riemannian manifold of the analytically intractable likelihood.

non-linear likelihood, computational expensive, proposals expensive, analytically intractable likelihood

@@

\subsection{Stochastic Population Models}

\subsection{Moment Closure Scheme}

@@

\subsection{Delay Acceptance Scheme}
